%!TEX root = thesis.tex

It has been a wonderful journey over the last four years. I am very lucky to have an overall extremely positive PhD experience, thanks to many people who have helped me along the way.

First of all, I want to express my deepest gratitude to my advisor Dan Ellis and co-advisor David Blei. Thanks Dan for allowing me a great deal of freedom to explore anything that I found interesting. Thanks Dave for taking me into his group when Dan was happily enjoying his sabbatical at Google -- I feel extremely fortunate, as I know how many people would like to have the opportunity to work with Dave. I learned a great deal from both of them about how to become a better researcher, and more importantly, a better person. I couldn't have possibly asked for better mentors along this journey.

I would also like to thank Shih-Fu Chang, John Paisley, and Matt Hoffman for serving on my thesis committee and providing valuable and insightful feedbacks. Special thanks to Matt, who also served as my mentor when I interned at Adobe Research. He is not only a good friend, but also has practically become a ``co-advisor'' to me during my early years in the PhD program. My research trajectory would have been completely different (probably in a bad way) if not with his guidance and encouragement. 

I am glad that I had LabROSA and Blei Lab as my intellectual home at Columbia. I will remember LabROSA as this room without a window but with a lot of fun, thanks to those who I have met and worked with over the years: Zhuo Chen, Colin Raffel, Brian McFee, Thierry Bertin-Mahieux, Courtenay Cotton, Rachel Kurtz, Minshu Zhan, Matt McVicar, Jiaming Liu, Cyril Gaudefroy, H\'{e}l\`{e}ne Papadopoulos, Diego Furtado Silva and others that I must have missed. I also thank all the folks in Blei Lab: Laurent Charlin, Jaan Altosaar, James McInerney, Maja Rudolph, Dustin Tran, Rajesh Ranganath, Alp Kucukelbir, Allison Chaney, Adji Dieng, Kriste Krstovski, Stephan Mandt, Francisco Ruiz, Kui Tang, and Yixin Wang. It's been a pleasure. I will remember all the insightful discussions and the eventual silences in the causal inference thematic meetings, as well as the late night beer (for me it was non-alcoholic) we had together, and I look forward to more in the future in the bars of NIPS/ICML/etc. Our NYU neighbor in the Music and Audio Research Lab also gave me a good reason to visit NYC downtown: Rachel Bittner, Eric Humphrey, Oriol Nieto and Justin Salamon, it has been great to know you all.

I would like to thank Roger Dannenberg and Guangyu Xia in the Compute Music Group at Carnegie Mellon University for getting me started my research on music information retrieval which led me to apply for a PhD program in the first place. I've already thanked them once in my master's thesis, so I will keep it short here.

Every summer, I did an internship and got to work with great mentors: Gautham Mysore, Erik Schmidt, and Keki Burjorjee, as well as many smart fellow interns, from whom I learned a lot. My experience at both Adobe Research and Pandora has been extremely rewarding thanks to all of them.

Being in a PhD program for a few years would be much more difficult without friends. Lucky for me, I got to be schoolmate again with one of my best friends from high school on the other side of the globe. I would like to thank Yanan Pei for listening to my rambling (or me listening to hers) about research and life, all the great (or not so great) food we have explored over the years, and many other things that would be so boring if I were to do them alone. Sorry that you will have to find someone else to have lunch/dinner with, I am sure you cannot find anyone better. Best of luck finishing up your PhD and hope our paths will cross again in the future.

Finally, I am grateful to my parents, Jianguo Liang and Jianmin Li, for everything. Not being able to be with you more often over the last few years hasn't been easy. Thank you for the unconditional support and love whenever I need them the most. Last but not least, I want to thank my wife Qimin Xu. Thank you for taking good care of me and being patient with me when I get cranky. It would have been significantly more difficult to survive graduate school were it not for your support every step of the way -- I look forward to our next adventure together. I don't say I love you very often but you know I do. 
